LCS Tool — An Algorithmic Introduction to Lagrangian Coherent Structures

### 2014-02-19 main.tex
Miscellaneous changes to tidy repository
Add step to strainline LCS algorithm to reflect remove_strain_in_shear.
Change shear LCS → elliptic LCS
Previously the terms shear and elliptic LCS were both used. Now only
elliptic LCS is used. Some of the LCS Tool functions and variables
still use the term shear however.

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage[uniquename=init,firstinits,backend=bibtex,sorting=none]{biblatex}
\usepackage{calc}
\usepackage{color}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{listings}
\usepackage[numbered,framed]{mcode}
\usepackage[space]{grffile}
\usepackage{lineno}
%% hyperref manual advises loading this package last
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,breaklinks=true}

@@

\linenumbers

@@

\subsection{Shear LCS}

Shear LCSs, or coherent Lagrangian vortices, are closed orbits in the Lagrangian shear vector fields $\boldsymbol \eta_{\pm}$ defined in Equation~\eqref{eq:etafields}. We find closed orbits by integrating lambda-lines starting from a Poincare section and evaluating the return map. A lambda-line returning to its starting point is a shear LCS. The main steps of the algorithm used to calculate shear LCS are enumerated in Table~\ref{t:Shear LCS algorithm} and described in detail below. The syntax of shear LCS functions in LCS Tool is in Table~\ref{t:Shear LCS functions}.

\begin{table}
\begin{center}
\begin{enumerate}
\item Position Poincare sections in flow domain to specify initial positions of lambda-lines
\item Integrate lambda-lines tangent to $\boldsymbol \eta_{\pm}$ (see Equation~\eqref{eq:etafields}.)
\item Calculate Poincare return map
\item Find closed orbits positions from Poincare return map
\item Identify outermost closed orbit
\end{enumerate}
\end{center}
\caption{Algorithm to calculate Shear LCS}
\label{t:Shear LCS algorithm}
\end{table}

\begin{table}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda)!}\\
\hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\\
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\\
\hline
\lstinline!lambda! & scalar lambda value in Equation~\eqref{eq:eta}\\
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![closedOrbits,orbits] = poincare_closed_orbit_multi(domain,resolution,shearline,PSList)!}\\
\hline
\lstinline!domain! & array to define flow domain\\
\hline
\lstinline!resolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\\
\hline
\lstinline!shearline! & structure of arrays of $\eta_+$ and $\eta_-$ values on main grid\\
\hline
\lstinline!PSList! & structure for Poincare section end-points, number of lambda-lines launched from Poincare section, and maximum closed lambda-line length\\
\hline
\lstinline!nBisection! & optional number of bisection steps to refine zero crossings of Poincare return map. Default is 5.\\
\hline
\lstinline!dThresh! & optional threshold to discard discontinuous zero crossings of Poincare return map. Default is $10^{-2}$.\\
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify lambda-line integration parameters\\
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default is \lstinline![false,false]!.\\
\hline \hline
\end{tabular}
\caption{LCS Tool shear LCS functions syntax}
\label{t:Shear LCS functions}
\end{table}

@@

An example of a Poincare return map produced from integration of the $\boldsymbol \eta_\pm$ field is shown in Figure~\ref{f:Poincare return map}. Well behaved orbits will return to the Poincare section and their integration will be stopped using ODE event detection. Some orbits may however deviate far from the Poincare section and their integration could take inordinately long. To limit this behavior, a maximum orbit length, \lstinline!poincareSection.orbitMaxLength!, is specified. In practice, viewing the Poincare section as the radius of a circle and setting the maximum lambda-line integration length to twice the circumference gives good results. In Figure~\ref{f:Poincare return map}, circle markers indicate closed orbit positions. Closed orbits exist where the Poincare return map, the distance between the final and the initial point of the orbit $P(s)-s$, is zero. The function \lstinline!poincare_closed_orbit_multi! performs the computations. In Figure~\ref{f:Poincare return map} not all zero crossings have circles. This is because a threshold parameter \lstinline!dThresh! is used to discard those zero crossings that appear in regions with high numerical noise. Its default value is 1\% of the length of the Poincare section. The position of each detected zero crossing is refined by iteratively bisecting the interval of the zero crossing. If after a predetermined number of iterations, \lstinline!nBisection! whose default value is five, the two points around the zero crossing have absolute values above \lstinline!dThresh!, the zero crossing is discarded. Once all valid closed lambda-line orbits have been located, the outermost orbit of every Poincare section is identified as a shear LCS.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/double_gyre/poincare_return_map}
\end{center}
\caption{Example of a Poincare return map obtained for shear LCS. Circle markers indicate closed orbit positions. The filled circle indicates the outermost closed orbit and shear LCS.}
\label{f:Poincare return map}
\end{figure}

@@

\begin{table}
\begin{enumerate}
\item Define a local maximization distance.
\item Find all local maxima of $\lambda_2$ on the main grid such that they are maxima within a circle with a radius equal to the local minimization distance.
\item Define a maximum strainline length
\item Integrate a strainline forward and backward according to Equation~\eqref{eq:strainline} and using the largest $\lambda_2$ local maximum as the initial position. Integrate until the strainline has attained the maximum strainline length, or until it has reached the domain boundary.
\item Flag any remaining local maxima of $\lambda_2$ within the minimization distance of the strainline as ineligible initial positions for subsequent strainlines
\item Continue integrating strainlines using local maxima of $\lambda_2$ as initial positions until no eligible local maxima of $\lambda_2$ remain.
\end{enumerate}
\caption{Algorithm to calculate strainline hyperbolic LCSs. Algorithm for stretchline hyperbolic LCSs is similar.}
\label{t:Hyperbolic LCS algorithm}
\end{table}

@@

\lstinputlisting[caption={Bickley jet script for hyperbolic and shear LCSs. Subset of \lstinline!demo/bickley_jet/hyperbolic_shear_lcs.m!},label=lst:Bickley jet hyperbolic_shear_lcs subset]{listings/bickley_jet/hyperbolic_shear_lcs_subset.m}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage[uniquename=init,firstinits,backend=bibtex,sorting=none]{biblatex}
\usepackage{calc}
\usepackage{color}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{listings}
\usepackage[numbered,framed]{mcode}
\usepackage[space]{grffile}
%% hyperref manual advises loading this package last
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,breaklinks=true}

@@

@@

\subsection{Elliptic LCS}

Elliptic LCSs, or coherent Lagrangian vortices, are closed orbits in the Lagrangian shear vector fields $\boldsymbol \eta_{\pm}$ defined in Equation~\eqref{eq:etafields}. We find closed orbits by integrating lambda-lines starting from a Poincare section and evaluating the return map. A lambda-line returning to its starting point is an elliptic LCS. The main steps of the algorithm used to calculate elliptic LCSs are enumerated in Table~\ref{t:Elliptic LCS algorithm} and described in detail below. The syntax of elliptic LCS functions in LCS Tool is in Table~\ref{t:Elliptic LCS functions}.

\begin{table}
\begin{center}
\begin{enumerate}
\item Position Poincare sections in flow domain to specify initial positions of lambda-lines
\item Integrate lambda-lines tangent to $\boldsymbol \eta_{\pm}$ (see Equation~\eqref{eq:etafields}.)
\item Calculate Poincare return map
\item Find closed orbits positions from Poincare return map
\item Identify outermost closed orbit
\end{enumerate}
\end{center}
\caption{Algorithm to calculate elliptic LCSs}
\label{t:Elliptic LCS algorithm}
\end{table}

\begin{table}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda)!}\\
\hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\\
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\\
\hline
\lstinline!lambda! & scalar lambda value in Equation~\eqref{eq:eta}\\
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![closedOrbits,orbits] = poincare_closed_orbit_multi(domain,resolution,shearline,PSList)!}\\
\hline
\lstinline!domain! & array to define flow domain\\
\hline
\lstinline!resolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\\
\hline
\lstinline!shearline! & structure of arrays of $\eta_+$ and $\eta_-$ values on main grid\\
\hline
\lstinline!PSList! & structure for Poincare section end-points, number of lambda-lines launched from Poincare section, and maximum closed lambda-line length\\
\hline
\lstinline!nBisection! & optional number of bisection steps to refine zero crossings of Poincare return map. Default is 5.\\
\hline
\lstinline!dThresh! & optional threshold to discard discontinuous zero crossings of Poincare return map. Default is $10^{-2}$.\\
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify lambda-line integration parameters\\
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default is \lstinline![false,false]!.\\
\hline \hline
\end{tabular}
\caption{LCS Tool elliptic LCS functions syntax}
\label{t:Elliptic LCS functions}
\end{table}

@@

An example of a Poincare return map produced from integration of the $\boldsymbol \eta_\pm$ field is shown in Figure~\ref{f:Poincare return map}. Well behaved orbits will return to the Poincare section and their integration will be stopped using ODE event detection. Some orbits may however deviate far from the Poincare section and their integration could take inordinately long. To limit this behavior, a maximum orbit length, \lstinline!poincareSection.orbitMaxLength!, is specified. In practice, viewing the Poincare section as the radius of a circle and setting the maximum lambda-line integration length to twice the circumference gives good results. In Figure~\ref{f:Poincare return map}, circle markers indicate closed orbit positions. Closed orbits exist where the Poincare return map, the distance between the final and the initial point of the orbit $P(s)-s$, is zero. The function \lstinline!poincare_closed_orbit_multi! performs the computations. In Figure~\ref{f:Poincare return map} not all zero crossings have circles. This is because a threshold parameter \lstinline!dThresh! is used to discard those zero crossings that appear in regions with high numerical noise. Its default value is 1\% of the length of the Poincare section. The position of each detected zero crossing is refined by iteratively bisecting the interval of the zero crossing. If after a predetermined number of iterations, \lstinline!nBisection! whose default value is five, the two points around the zero crossing have absolute values above \lstinline!dThresh!, the zero crossing is discarded. Once all valid closed lambda-line orbits have been located, the outermost orbit of every Poincare section is identified as an elliptic LCS.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/double_gyre/poincare_return_map}
\end{center}
\caption{Example of a Poincare return map obtained for an elliptic LCS. Circle markers indicate closed orbit positions. The filled circle indicates the outermost closed orbit and elliptic LCS.}
\label{f:Poincare return map}
\end{figure}

@@

\begin{table}
\begin{enumerate}
\item Define a local maximization distance.
\item Find all local maxima of $\lambda_2$ on the main grid such that they are maxima within a circle with a radius equal to the local minimization distance.
\item Define a maximum strainline length
\item Integrate a strainline forward and backward according to Equation~\eqref{eq:strainline} and using the largest $\lambda_2$ local maximum as the initial position. Integrate until the strainline has attained the maximum strainline length, or until it has reached the domain boundary.
\item Flag any remaining local maxima of $\lambda_2$ within the minimization distance of the strainline as ineligible initial positions for subsequent strainlines
\item Continue integrating strainlines using local maxima of $\lambda_2$ as initial positions until no eligible local maxima of $\lambda_2$ remain.
\item Remove all strainline segments within elliptic LCSs.
\end{enumerate}
\caption{Algorithm to calculate strainline hyperbolic LCSs. Algorithm for stretchline hyperbolic LCSs is similar.}
\label{t:Hyperbolic LCS algorithm}
\end{table}

@@

\lstinputlisting[caption={Bickley jet script for hyperbolic and elliptic LCSs. Subset of \lstinline!demo/bickley_jet/hyperbolic_shear_lcs.m!},label=lst:Bickley jet hyperbolic_shear_lcs subset]{listings/bickley_jet/hyperbolic_shear_lcs_subset.m}
